---
title: "LC-MS/MS Raw Spectral Data Processing"
author: "Debojyoti"
date: "2025-02-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(MetaboAnalystR)
library(OptiLCMS)
library(tidyr)
library(dplyr)
library(comprehenr)

```

```{r}
download.file("https://api2.xialab.ca/api/download/metaboanalyst/malaria_raw.zip",
    destfile = "malaria_raw.zip",
    method = "wget")
unzip("malaria_raw.zip", exdir = "upload")
```


```{r}
# Here, we extract ROIs from 3 QC samples.
DataFiles <- list.files("upload/malaria_LCMS/", pattern = "QC_\\d{3}.mzML", full.names = TRUE)

# Region of Interest (ROI) extraction
mSet <- MetaboAnalystR::PerformROIExtraction(datapath = DataFiles, rt.idx = 0.9, rmConts = TRUE)
```


# Auto-optimization of parameters

# Here we use PerformParamsOptimization to optimize parameters based on 
# the extracted ROI (stored in 'mSet') before process the entire dataset
```{r}

best_params <- MetaboAnalystR::PerformParamsOptimization(mSet, param = SetPeakParam(platform = "UPLC-Q/E"), ncore = 4)


get_group <- function(s){
    part1 <- strsplit(s,"/")[[1]][3]
    part2 <- strsplit(part1,"_")[[1]][1]
    return(part2)
}

sample_name <- list.files("upload/malaria_LCMS", full.names = TRUE, pattern = "mzML")
meta_data <- data.frame(sample_name=sample_name) %>%
    dplyr::mutate(group_name = sapply(sample_name, get_group))

meta_data_to_use <- meta_data %>% 
    dplyr::mutate(sample_name = sub(".mzML", "", basename(sample_name)))

# "path" is used to specify the path to the folder containing the raw MS spectra to be processed.
# BPI and TIC plotting can be enabled with parameter, 
# "plotSettings = SetPlotParam(Plot = T)", or disabled by changing "T" into "F";

mSet <- OptiLCMS::ImportRawMSData(
    path = meta_data$sample_name,
    metadata = meta_data_to_use, 
    plotSettings = SetPlotParam(Plot = T)
    )
```


# Raw spectral data processing
# "mSet" include complete raw MS spectra to be processed.
# "params" is using the "best_params" generated above
# Plotting functions can be enabled with parameter, 
# "plotSettings = SetPlotParam(Plot = T)", or disabled by changing "T" into "F";
```{r}
mSet <- MetaboAnalystR::PerformPeakProfiling(
    mSet, 
    Params = best_params, 
    plotSettings = SetPlotParam(Plot=TRUE)
    )
```


# Feature annotation
# We firstly define the parameters for feature annotation

# 'polarity' is required, can be either 'negative' or 'positive';
# 'perc_fwhm' is used to set the percentage of the width of the FWHM for peak grouping. 
#              Default is set to 0.6;
# 'mz_abs_iso' is used to set the allowed variance for the search (for isotope annotation). 
#              The default is set to 0.005;
# 'max_charge' is set the maximum number of the isotope charge. 
#              For example, the default is 2, therefore the max isotope charge is 2+/-;
# 'max_iso' is used to set the maximum number of isotope peaks.
#              For example, the default is 2, therefore the max number of isotopes per peaks is 2;
# 'corr_eic_th' is used to set the threshold for intensity correlations across samples. 
#              Default is set to 0.85.
# 'mz_abs_add' is used to set the allowed variance for the search (for adduct annotation). 
#              Default is set to 0.001.
# 'adducts' is used to specify the adducts based on your instrument settings.
```{r}
annParams <-OptiLCMS::SetAnnotationParam(polarity = 'positive', mz_abs_add = 0.015);

# "mSet" include processed raw MS spectra to be processed.
# "annParams" is the parameters used for annotation

mSet <- OptiLCMS::PerformPeakAnnotation(mSet, annParams)

```


# Feature table generation

# Here we format and filter the peak list for following analysis with MetaboAnalystR

# Parameters are explained as below,
# annParams, is the object created using the SetAnnotationParam function above;
# filtIso, is used to decide to filter out all isotopes (TRUE) or not (FALSE);
# filtAdducts, is used to decide to filter out all adducts (TRUE) or not (FALSE);
# missPercent, specify the threshold to remove features missing in a certain percentage
#              of all samples in a group.
```{r}
mSet <- OptiLCMS::FormatPeakList(
    mSet, 
    annParams, 
    filtIso = FALSE, 
    filtAdducts = FALSE, 
    missPercent = 1)
```

Export annotation results, the annotation will be save as "annotated_peaklist.csv"
```{r}
OptiLCMS::Export.Annotation(mSet)
```

Export complete feature table. It will be saved as "metaboanalyst_input.csv"
This table can be used for statistic analysis, functional analysis, biomarker analysis module directly.

```{r}
OptiLCMS::Export.PeakTable(mSet)
```

Export a summary table (peak_result_summary.txt) to summarize the information of all peaks in
different samples. The columns are sample names, groups, retention time range, m/z range of all peaks,
number of all peaks and percentage of missing features.
```{r}

OptiLCMS::Export.PeakSummary(mSet)
```
